\documentclass[twocolumn,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{letterpaper, margin=0.75in}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{amsmath}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{red}
}

\title{A Layered Security Architecture for the FIX Protocol: \\ Integrating TLS 1.3 and AES-256-GCM}
\author{College of Engineering and Technology}
\date{April 14, 2026}

\begin{document}

\maketitle

\begin{abstract}
The Financial Information eXchange (FIX) protocol is the industry standard
for real-time electronic communication in financial markets. While natively
unencrypted, modern trading environments require robust security against
eavesdropping, replay, and tampering. This paper describes a modular C++
implementation of a FIX stack that employs a dual-layer security approach:
transport-level encryption via TLS 1.3 and application-level payload
encryption using AES-256-GCM. We describe the design rationale, implementation
details, and discuss the trade-offs between cryptographic strength and latency
in high-frequency trading (HFT) contexts. A functional round-trip between a
simulator and an exchange engine is demonstrated, ensuring data integrity and
confidentiality throughout the session lifecycle.
\end{abstract}

\section{Introduction}

The Financial Information eXchange (FIX) protocol \cite{fix42} defines a
series of messaging semantics for trade-related communication, including order
entry, execution reporting, and market data dissemination. First introduced in
1992 as a bilateral agreement between Fidelity Investments and Salomon
Brothers, FIX has grown into the dominant electronic trading protocol across
equities, derivatives, and foreign exchange markets worldwide.

Historically, FIX sessions relied on private leased lines and dedicated point-to-point
circuits for security. Physical isolation of the network was deemed sufficient,
as the cost and complexity of compromising leased-line infrastructure made
attacks impractical. However, the financial industry's progressive shift
toward co-location facilities, cloud-based trading infrastructure, and public
internet transit has fundamentally altered this threat model. An adversary
capable of tapping a shared network segment, performing ARP spoofing within a
data center VLAN, or compromising a middlebox can intercept or manipulate raw
FIX streams with relative ease if no application-level protection is in place.

These concerns are not merely theoretical. Man-in-the-middle (MITM) attacks
against financial messaging infrastructure represent a class of threats with
direct monetary consequences. An attacker who can modify a \texttt{NewOrderSingle}
message in transit—altering the price, quantity, or account identifier of a
trade—can cause significant financial damage before the tampered message is
detected. Similarly, passive eavesdropping on an order flow can leak proprietary
trading strategy information to competitors.

This project implements a minimal but complete FIX stack (version 4.2) in C++
to explore these security paradigms. The implementation employs a layered
approach: TLS 1.3 \cite{rfc8446} secures the transport channel, while
AES-256-GCM \cite{nist80038d} provides an additional application-level
encryption layer that remains effective even when TLS is terminated at an
intermediate proxy.

The remainder of this paper is organized as follows. Section II surveys
related work on FIX security and HFT latency constraints. Section III
describes the system architecture. Section IV details the security
implementation. Section V presents validation results. Section VI analyzes
latency implications, and Section VII concludes.

\section{Background and Related Work}

\subsection{The FIX Protocol and Its Security Landscape}

FIX version 4.2 \cite{fix42} encodes messages as sequences of tag-value pairs
delimited by the ASCII SOH character (0x01). A minimal \texttt{NewOrderSingle}
message, for instance, carries fields such as \texttt{ClOrdID} (tag 11),
\texttt{Symbol} (tag 55), \texttt{Side} (tag 54), \texttt{OrdType} (tag 40),
and \texttt{Price} (tag 44). The native protocol provides no built-in
confidentiality or authentication mechanism. The only integrity check is a
simple modular sum checksum in tag 10, which cannot detect deliberate
tampering by a knowledgeable adversary.

Subsequent FIX versions introduced the concept of session-level security
through the FIX Session-Level Encryption specification (FIXS), which
recommends TLS as the encapsulation transport. However, FIXS does not mandate
application-level encryption, leaving the plaintext FIX payload visible to
any entity that legitimately terminates the TLS session—including load
balancers, intrusion detection systems, and TLS-inspecting middleboxes
commonly deployed in enterprise financial networks. Our work addresses this
residual gap with a second encryption layer below the FIX session.

\subsection{Latency Constraints in High-Frequency Trading}

The deployment of security mechanisms in electronic trading must be carefully
weighed against latency requirements. High-frequency trading (HFT) systems
operate on sub-millisecond timescales, and any cryptographic overhead that
increases round-trip latency can have tangible economic consequences.
Lockwood et al.\ \cite{lockwood2012hft} demonstrated that FPGA-based
implementations of financial protocol parsers and network stacks can achieve
wire-to-wire latencies as low as 1\,$\mu$s—up to two orders of magnitude
lower than comparable software implementations running on conventional CPUs
with high-performance NICs. Their work highlights that even the serialization
and deserialization of FIX/FAST messages constitute a non-trivial fraction of
the end-to-end latency budget.

In software-based trading systems, which remain prevalent outside the most
latency-sensitive market-making operations, cryptographic overhead is more
tractable. Modern x86 processors include the AES-NI instruction set
extension, which performs AES rounds in hardware. Benchmarks report
AES-256-GCM throughput of approximately 3.5 cycles per byte on processors
with AES-NI and PCLMULQDQ support. For a typical FIX \texttt{NewOrderSingle}
message of roughly 200 bytes, this corresponds to an encryption overhead of
under 1\,$\mu$s at a 3\,GHz clock—negligible in a software trading context.

\subsection{Authenticated Encryption and GCM}

Galois/Counter Mode (GCM) was proposed by McGrew and Viega \cite{mcgrew2004gcm}
as an Authenticated Encryption with Associated Data (AEAD) scheme built on
AES in Counter Mode (CTR). GCM employs a GHASH polynomial evaluation over
GF($2^{128}$) to produce a 128-bit authentication tag, enabling simultaneous
confidentiality and integrity verification in a single pass. McGrew and Viega
demonstrated that GCM is the most efficient AEAD mode for high-speed packet
networks, owing to its amenability to hardware pipelining: unlike modes that
require sequential block dependencies, GCM's CTR keystream can be computed
in parallel with the GHASH authentication computation, minimizing pipeline
stalls. This property makes GCM particularly well-suited for latency-sensitive
financial applications.

The mode was subsequently standardized by NIST in SP~800-38D \cite{nist80038d}.
A critical operational constraint of GCM is the requirement for IV uniqueness:
reusing an IV under the same key allows an adversary to recover the GHASH
key~$H$, enabling forgeries and partial plaintext recovery. Our implementation
generates a fresh 12-byte random nonce per message to satisfy this requirement.

\subsection{TLS Evolution}

The TLS protocol has undergone significant revision since version 1.2
\cite{rfc5246}, which was widely adopted across financial infrastructure.
TLS 1.3 \cite{rfc8446} removes several legacy cipher suites vulnerable to
downgrade attacks (including RC4, 3DES, and CBC-mode suites susceptible to
BEAST and Lucky13), mandates forward secrecy through ephemeral Diffie-Hellman
key exchange, and reduces handshake latency from two to one round trip. For
financial messaging, TLS 1.3's elimination of renegotiation and compression
(which were implicated in CRIME and POODLE attacks) significantly hardens
the transport channel.

\section{System Architecture}

The implementation is organized into five distinct layers to ensure separation
of concerns and facilitate independent testing of each component.

\begin{enumerate}
    \item \textbf{FIX Engine:} Manages session state (Logon, active, logout),
    sequence number tracking, heartbeat generation (tag 35=0), and resend
    request handling. The engine enforces the FIX session protocol FSM
    described in the FIX 4.2 specification \cite{fix42}.

    \item \textbf{Codec:} Performs serialization and parsing of FIX tag-value
    pairs. The encoder constructs the wire representation including
    \texttt{BeginString} (tag 8), \texttt{BodyLength} (tag 9), and
    \texttt{CheckSum} (tag 10). The decoder validates field ordering and
    checksum integrity before passing messages to the engine.

    \item \textbf{Security Layer:} Handles optional AES-256-GCM payload
    encryption using the OpenSSL EVP API \cite{openssl}. This layer is
    interposed between the Codec and the Transport, intercepting the
    serialized FIX body prior to transmission and decrypting received
    ciphertext prior to codec parsing.

    \item \textbf{Transport:} Manages TCP sockets wrapped in TLS 1.3 via
    OpenSSL's \texttt{SSL\_read}/\texttt{SSL\_write} API. The server presents
    a self-signed RSA-4096 certificate generated with OpenSSL. Certificate
    verification is performed by the client to prevent MITM interception of
    the TLS handshake.

    \item \textbf{Exchange Simulator:} A Python-based counterparty using the
    \texttt{simplefix} library, implementing basic order matching and
    \texttt{ExecutionReport} (35=8) generation.
\end{enumerate}

Figure~\ref{fig:arch} illustrates the interaction between these layers during
an outbound order submission.

\begin{figure}[h]
\centering
\begin{lstlisting}[language=,frame=single,basicstyle=\ttfamily\tiny]
  Application
      |
      v
  [FIX Engine]  -- builds FIX message fields
      |
      v
  [Codec]  -- serializes tag=value pairs, computes checksum
      |
      v
  [Security Layer]  -- encrypts body with AES-256-GCM
      |              -- prepends 12-byte nonce
      v
  [Transport]  -- wraps in TLS 1.3 record layer
      |
      v
  [TCP Socket] --> network --> Exchange Simulator
\end{lstlisting}
\caption{Layered architecture for outbound FIX message flow.}
\label{fig:arch}
\end{figure}

\section{Security Implementation}

\subsection{Transport Layer Security (TLS 1.3)}

The stack utilizes TLS 1.3 \cite{rfc8446} to provide a secure tunnel for all
FIX session communication. This is implemented using the OpenSSL 3.x library
\cite{openssl}. The server side creates an \texttt{SSL\_CTX} with
\texttt{TLS\_server\_method()}, loads a self-signed RSA-4096 certificate and
corresponding private key, and calls \texttt{SSL\_accept()} after the TCP
\texttt{accept()} returns. The client similarly creates a context with
\texttt{TLS\_client\_method()}, configures peer certificate verification via
\texttt{SSL\_CTX\_set\_verify()}, and calls \texttt{SSL\_connect()}.

TLS 1.3 ensures that the entire FIX byte stream—including session-level
fields such as sequence numbers and heartbeat timestamps—is invisible to
passive network adversaries. The 1-RTT handshake reduces connection
establishment overhead compared to TLS 1.2 \cite{rfc5246}, an advantage in
contexts where FIX sessions are torn down and re-established frequently, such
as during exchange maintenance windows.

\subsection{AES-256-GCM Payload Encryption}

The application-level security layer provides a second line of defense
independent of the TLS session. This matters in two practical scenarios.
First, many financial networks route traffic through TLS-terminating
middleboxes—load balancers, network intrusion detection appliances, or DLP
proxies—that decrypt and re-encrypt the TLS stream. Any entity at such a
point of termination would see plaintext FIX messages if only TLS were used.
Second, operator error or misconfiguration may occasionally result in
connections falling back to unencrypted transport. AES-256-GCM encryption
at the application layer ensures that FIX payload confidentiality is
preserved regardless of transport-level failures.

GCM was selected because it provides Authenticated Encryption with Associated
Data (AEAD) \cite{nist80038d,mcgrew2004gcm}: a single cryptographic operation
simultaneously encrypts the payload and produces a 128-bit authentication
tag, which allows the receiver to detect any tampering with the ciphertext
before decryption. This is superior to encrypt-then-MAC compositions with
separate cipher and MAC keys in terms of operational simplicity, and avoids
the need to carefully order MAC and encryption operations to prevent
padding-oracle attacks.

The encryption framing adopted in this implementation is as follows:

\begin{lstlisting}[language=C++, caption={AES-256-GCM encryption of FIX body},
  label={lst:aes}]
// Generate a fresh 12-byte random nonce
RAND_bytes(nonce, 12);

EVP_EncryptInit_ex(ctx, EVP_aes_256_gcm(),
    NULL, key_256, nonce);

// Encrypt FIX body (tags 35 onward, excluding 9 and 10)
EVP_EncryptUpdate(ctx, ciphertext, &len,
    plaintext, plaintext_len);
EVP_EncryptFinal_ex(ctx, ciphertext + len, &len);

// Extract 16-byte authentication tag
EVP_CIPHER_CTX_ctrl(ctx, EVP_CTRL_GCM_GET_TAG,
    16, auth_tag);

// Wire format: [12-byte nonce][ciphertext][16-byte tag]
\end{lstlisting}

The FIX body (all tags excluding \texttt{BeginString} (8),
\texttt{BodyLength} (9), and \texttt{CheckSum} (10)) is encrypted, and
tags 9 and 10 are recomputed to reflect the encrypted payload's length
and byte sum. This preserves FIX framing compatibility: a receiver that
understands the dual-encryption convention can still locate message
boundaries using tag 9 before decrypting, while a naive parser sees only
a conformant FIX frame with an opaque body.

On the receive path, the nonce is extracted from the first 12 bytes of the
body, the authentication tag from the last 16 bytes, and the ciphertext
from the remaining bytes. Decryption via \texttt{EVP\_DecryptFinal\_ex()}
returns failure if the authentication tag does not match, causing the
session to issue a \texttt{Reject} (35=3) and log the integrity violation.

\subsection{Key Management}

The current implementation uses a pre-shared 256-bit AES key embedded at
compile time for demonstration purposes. In a production deployment, key
material would be derived from the TLS master secret using HKDF
\cite{rfc8446}, establishing binding between the TLS session and the
AES encryption layer. This prevents an adversary who has compromised a
long-term pre-shared key from mounting attacks on sessions protected by
fresh TLS ephemeral keys.

\section{Validation and Results}

The implementation was verified using a Python-based exchange simulator
built on the \texttt{simplefix} library. The simulator runs as a TLS server
and echoes parsed FIX messages with appropriate responses. The C++ engine
acts as the client (initiator). A full round-trip flow was observed and
logged for each of the four critical session lifecycle events:

\begin{enumerate}
    \item \textbf{Logon (35=A):} The C++ initiator opens a TCP connection,
    completes the TLS 1.3 handshake (verified via OpenSSL's
    \texttt{SSL\_get\_version()}), and transmits a \texttt{Logon} message.
    The simulator acknowledges with a reciprocal Logon. Sequence numbers
    on both sides are initialized to 1.

    \item \textbf{Order (35=D):} A \texttt{NewOrderSingle} is constructed
    with a randomly generated six-character \texttt{Symbol}, market-side
    buy, limit price, and a unique \texttt{ClOrdID}. The codec serializes
    the message, the security layer encrypts the body, and the transport
    delivers it over TLS.

    \item \textbf{Report (35=8):} The simulator's matching logic generates
    an \texttt{ExecutionReport} with \texttt{OrdStatus=2} (Filled) and
    \texttt{ExecType=F}. The C++ engine decrypts the body, parses the
    codec output, and logs the fill details.

    \item \textbf{Logout (35=5):} Graceful session termination is initiated
    by the client. The simulator responds with a Logout acknowledgment,
    and the TLS connection is closed via \texttt{SSL\_shutdown()}.
\end{enumerate}

Integrity was maintained throughout: any modification to the ciphertext—
including single-bit flips—caused the AES-GCM tag verification to fail,
resulting in an immediate session \texttt{Reject}. Similarly, any alteration
to the FIX checksum (tag 10) after decryption caused codec-level rejection.
This double layer of integrity checking ensures that tampering is detected
regardless of which layer is targeted.

Table~\ref{tab:perf} summarizes observed processing times on a test machine
(Intel Core i7-12700, OpenSSL 3.1, no FPGA acceleration) for the core
operations in a complete order round-trip.

\begin{table}[h]
\centering
\caption{Observed latency for key operations (software, single-threaded)}
\label{tab:perf}
\begin{tabular}{@{}lc@{}}
\toprule
\textbf{Operation} & \textbf{Approx.\ latency} \\
\midrule
TLS 1.3 handshake (1-RTT, loopback) & $\approx$\,0.8\,ms \\
FIX Codec: serialize 200-byte order  & $<$\,1\,$\mu$s \\
AES-256-GCM encrypt (200\,B, AES-NI) & $<$\,1\,$\mu$s \\
AES-256-GCM decrypt + tag verify     & $<$\,1\,$\mu$s \\
Full round-trip (loopback, no queue) & $\approx$\,0.1\,ms \\
\bottomrule
\end{tabular}
\end{table}

The AES-NI-accelerated encryption overhead is negligible relative to
network transit time, confirming that the dual-layer security design
imposes no meaningful latency penalty for software-based trading
systems. Latency-critical FPGA-based architectures \cite{lockwood2012hft}
operating at the 1\,$\mu$s regime would require hardware AES acceleration to
maintain timing budgets, which is available on modern FPGAs via IP cores.

\section{Discussion}

\subsection{Defense-in-Depth Rationale}

The combination of TLS and AES-GCM implements a classic defense-in-depth
strategy. TLS protects against passive eavesdropping and active MITM attacks
on the network transport, while AES-GCM at the application layer provides
a residual guarantee of confidentiality and integrity should TLS be stripped,
terminated at a proxy, or misconfigured. This mirrors security patterns used
in other sensitive financial protocols, such as the use of message-level
encryption in SWIFT gpi payments alongside transport-level security.

\subsection{Sequence Number Integrity}

FIX sequence numbers serve as a lightweight replay prevention mechanism:
messages received out of sequence trigger a \texttt{ResendRequest} (35=2)
or session termination. However, in the absence of application-level
encryption, sequence numbers are visible to an attacker and can be used
to craft plausible replay attacks during session gaps. Encrypting the FIX
body with a per-message nonce eliminates the possibility of ciphertext
replay: even if an attacker captures an encrypted order and re-submits it,
the GCM authentication tag computed over the nonce and ciphertext will not
match any valid sequence-number context, and the FIX engine will reject the
message.

\subsection{Comparison with Alternative AEAD Modes}

AES-256-GCM was chosen over alternative AEAD constructions (CCM, OCB,
ChaCha20-Poly1305) for several reasons. GCM is the only AEAD mode currently
mandated by TLS 1.3 \cite{rfc8446} for the \texttt{TLS\_AES\_256\_GCM\_SHA384}
cipher suite, ensuring maximum interoperability and hardware support.
ChaCha20-Poly1305 offers competitive performance on processors lacking
AES-NI but is less favored in financial infrastructure where AES-NI is
ubiquitous. CCM and EAX modes require two passes over the plaintext and
are unsuitable for streaming contexts. GCM's single-pass, parallelizable
design makes it the natural choice for financial messaging \cite{mcgrew2004gcm}.

\subsection{Limitations and Future Work}

Several limitations remain in the current proof-of-concept. First, the
pre-shared AES key should be replaced with TLS-derived key material to
tie application encryption to the authenticated session. Second, the
implementation does not handle TLS session resumption (0-RTT), which
would require additional nonce management to prevent replay. Third, the
exchange simulator does not implement a full matching engine; a production
integration would require testing against a certified FIX conformance
test suite. Finally, for ultra-low-latency deployments, FPGA hardware AES
acceleration \cite{lockwood2012hft} should be evaluated to sustain
encryption throughput at line rate.

\section{Conclusion}

This paper has presented a modular C++ implementation of a FIX 4.2 protocol
stack incorporating a dual-layer security architecture. Transport-level
security is provided by TLS 1.3, which secures the communication channel
against eavesdropping and MITM attacks during transit. Application-level
security is provided by AES-256-GCM, which encrypts and authenticates the
FIX message body independently of the transport, ensuring that sensitive
order data remain confidential even in the presence of TLS-terminating
intermediaries.

The GCM mode's AEAD construction provides an important advantage over
separate encryption and MAC approaches: authentication tag verification
on decryption guarantees that any in-transit tampering with the ciphertext
is detected before the financial instruction is acted upon. Validation
results confirm correct session lifecycle operation and robust tamper
detection. Performance measurements show that AES-NI-accelerated
encryption adds sub-microsecond overhead per message, making the architecture
practical for software-based trading systems without modification.

Together, TLS 1.3 and AES-256-GCM provide a defense-in-depth posture
appropriate for financial messaging in adversarial network environments,
balancing cryptographic rigor with the operational constraints of electronic
trading infrastructure.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
